P.berghei pir analysis

### 2021-03-12 Manuscript_pirAnalysis_Github/Pberghei/Manuscript_Github_Pberghei_pir.Rmd
Updated names, orders, to match the manuscript. Removed figures.

pir_clade <- mutate(pir_clade, 
                    locus = pir_loci$locus[match(Geneid, pir_loci$Geneid)])

pir_clade <- read_delim('../standard_files/Pberghei_pir_clades.txt', 
                        delim = '\t')
pir_id <- pir_clade$Geneid

@@

## Supp figure - PCA of all the samples.

How many experiments cover each life stage?

```{r experiment_per_life_stage}

big_tpm %>% 
  rownames_to_column(var = 'Geneid') %>% 
  melt(id.vars = 'Geneid') %>% 
  mutate(
    stage = str_extract(variable, pattern = '(?<=E[:digit:]{2}_).+(?=_[:digit:])'),
    expt = str_extract(variable, pattern = 'E[:digit:]{2}')
  ) %>% 
  group_by(stage) %>% 
    summarise(num_expt = n_distinct(expt)) %>% 
    arrange(desc(num_expt))

@@

###

### Read in top genes

###

@@

## Pearson correlation plot of collated samples.

```{r all_samples_collated_heatmap}
## Collated across the stages for a more easily understood heatmap

@@

@@

```{r SorL_stacked_eachGene_fig1b}
  tpm_data <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    filter(Geneid %in% pir_id) %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
         ) %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
    mutate(stack_group = Geneid) %>% 
    mutate(tpm = ifelse(tpm > 1, tpm, 0)) %>% #Filtering step for low value pir genes.
    group_by(stage, 
             stack_group,
             clade) %>% 
    dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(clade),
          total_tpm
)

@@

## Figure 1C

@@

#Separate out the pan-rodent pir
# tpm_data_subfam$clade[tpm_data_subfam$stack_group == 'PBANKA_0524600'] <- 'pan-rodent'

@@

## Chromosome proportional bar chart supp fig

```{r chrom_prop_bar_chart}
tpm_data_chrom <- combined_data %>% 
  rownames_to_column(var = 'Geneid') %>% 
  filter(Geneid %in% pir_id) %>% 
  melt(id.vars = 'Geneid', 
       variable.name = 'stage', 
       value.name = 'tpm') %>% 
  mutate(
    chrom = str_extract(Geneid, pattern = '(?<=PBANKA_)[[:digit:]]{2}')
  ) %>% 
  group_by(chrom) %>% 
  mutate(chrom = paste0(chrom, ' (n = ', n_distinct(Geneid) ,')')) %>% 
  group_by(stage, 
           chrom) %>% 
  dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(chrom),
          total_tpm
) %>% 
  mutate(chrom = factor(chrom, 
                        levels = unique(chrom)))

  stacked_max <- tpm_data_chrom %>% 
    ungroup %>% 
    group_by(stage) %>% 
    dplyr::summarise(max_tpm = sum(total_tpm)) %>% 
    .$max_tpm %>% 
    max

y <- ggplot(data = tpm_data_chrom) +
  geom_bar_pattern(
    aes(
      x = stage,
      y = total_tpm, 
      fill = chrom,
      group = stage,
      pattern = chrom
    ),
    position = 'fill',
    stat = 'identity',
    colour = 'black',
    size = 0.1,
    pattern_density = 0.2,
    pattern_spacing = 0.05,
    pattern_key_scale_factor = 0.25
  ) +
  scale_pattern_manual(
    values = rep(c('none', 
                   'circle'), 
                 each = length(c(cbPalette, 
                                 'red', 
                                 'white')), 
                 length.out = length(unique(tpm_data_chrom$chrom))),
    labels = unique(tpm_data_chrom$chrom),
    guide = guide_legend(reverse = TRUE),
    name = 'chromosome'
  ) +
  scale_fill_manual(
    breaks = unique(tpm_data_chrom$chrom),
    values = rep_len(c(cbPalette, 
                       'red', 
                       'white'),
                     length.out = length(unique(tpm_data_chrom$chrom))),
    labels =  unique(tpm_data_chrom$chrom),
    guide = guide_legend(reverse = TRUE),
    name = 'chromosome'
  ) +
  scale_color_manual('black') +
  scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) +
  facet_grid(.~stage,
             scales = "free_x",
             space = "free_x") +
  scale_x_discrete(expand=expand_scale(add=1)) +
  theme_classic() +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    strip.text.x = element_text(size=8,
                                face = 'bold',
                                angle = 90),
    strip.background = element_rect(
      color="black",
      size=0.1,
      linetype="solid"
    ),
    panel.spacing.x = unit(0, "null")
  )

    levels(tpm_data$stage)
  )
  fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                     str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                     str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                     str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
  k <- 1
  for (i in strip_t) {
    j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
    g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
    k <- k+1
  }

## Expression by locus. Proportion. Supp figure.

Loci are defined by L or R of each chromosome number, will define this by the jump in the gene numbers.

```{r defining_loci}

tpm_data_locus <- combined_data %>% 
  rownames_to_column(var = 'Geneid') %>% 
  filter(Geneid %in% pir_id) %>% 
  melt(id.vars = 'Geneid', 
       variable.name = 'stage', 
       value.name = 'tpm') %>% 
  mutate(
    locus = pir_loci$locus[match(Geneid, pir_loci$Geneid)]
  ) %>% 
  filter(!is.na(locus)) %>% 
  group_by(locus) %>% 
  mutate(locus = paste0(locus, ' (n = ', n_distinct(Geneid) ,')')) %>% 
  group_by(stage, 
           locus) %>% 
  dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(locus),
          total_tpm
) %>% 
  mutate(locus = factor(locus, levels = unique(locus)))

  stacked_max <- tpm_data_locus %>% 
    ungroup %>% 
    group_by(stage) %>% 
    dplyr::summarise(max_tpm = sum(total_tpm)) %>% 
    .$max_tpm %>% 
    max

(y <- ggplot(data = tpm_data_locus) +
  geom_bar_pattern(
    aes(
      x = stage,
      y = total_tpm, 
      fill = locus,
      group = stage,
      pattern = locus
    ),
    position = 'fill',
    stat = 'identity',
    colour = 'black',
    size = 0.1,
        pattern_density = 0.05,
    pattern_spacing = 0.1,
    pattern_key_scale_factor = 0.1
    #  pattern_density = 0.1,
    # pattern_spacing = 0.1,
    # pattern_key_scale_factor = 0.5
  ) +
  scale_pattern_manual(
    values = rep(c('none', 
                   'circle',
                   'stripe'), 
                 each = length(c(cbPalette, 
                                 'red', 
                                 'white')), 
                 length.out = length(unique(tpm_data_locus$locus))),
    labels = unique(tpm_data_locus$locus),
    guide = guide_legend(reverse = TRUE),
    name = 'locus'
  ) +
  scale_fill_manual(
    breaks = unique(tpm_data_locus$locus),
    values = rep_len(c(cbPalette, 
                       'red', 
                       'white'),
                     length.out = length(unique(tpm_data_locus$locus))),
    labels =  unique(tpm_data_locus$locus),
    guide = guide_legend(reverse = TRUE),
    name = 'locus'
  ) +
  scale_color_manual('black') +
  scale_size(range = c(0,3)) +
  ylab(
    bquote('total '~italic(.('pir'))~' TPM')
  ) +
  facet_grid(.~stage,
             scales = "free_x",
             space = "free_x") +
  scale_x_discrete(expand=expand_scale(add=1)) +
  theme_classic() +
  theme(
    axis.title.x=element_blank(),
    axis.text.x=element_blank(),
    axis.ticks.x = element_blank(),
    # axis.line.x = element_blank(),
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    legend.key.size = unit(3,'mm'),
    legend.key.width = unit(2,'mm'),
    strip.text.x = element_text(size=8,
                                face = 'bold',
                                angle = 90),
    strip.background = element_rect(
      color="black",
      size=0.1,
      linetype="solid"
    ),
    panel.spacing.x = unit(0, "null")
  ))

    levels(tpm_data$stage)
  )
  fills <- case_when(str_detect(stage_vec, pattern = 'Asex') ~ 'coral1' ,
                     str_detect(stage_vec, pattern = 'Liv') ~ 'plum2',
                     str_detect(stage_vec, pattern = 'Gam') ~ 'lightskyblue1',
                     str_detect(stage_vec, pattern  = 'Bld|Sporo|Ook') ~ 'lightgreen')
  k <- 1
  for (i in strip_t) {
    j <- which(grepl('rect', g$grobs[[i]]$grobs[[1]]$childrenOrder))
    g$grobs[[i]]$grobs[[1]]$children[[j]]$gp$fill <- fills[k]
    k <- k+1
  }

  grid.newpage()
  grid.draw(g)
```
## Gametocyte marker gene expression - supp figure

>>>>

pir_clade <- read_delim('../standard_files/Pberghei_pir_clades.txt', 
                        delim = '\t')
pir_clade <- mutate(pir_clade, 
                    locus = pir_loci$locus[match(Geneid, pir_loci$Geneid)])
pir_id <- pir_clade$Geneid

@@

@@

@@

##Supp figure 1A - Pearson correlation plot of collated samples.

```{r all_samples_collated_heatmap}

## Collated across the stages for a more easily understood heatmap

@@

## Supp figure 1B - PCA of all the samples.

#Supp figure 1C - PCA using just the pir genes.

@@

```{r SorL_stacked_eachGene_fig2a}
  tpm_data <- combined_data %>% 
    rownames_to_column(var = 'Geneid') %>% 
    filter(Geneid %in% pir_id) %>% 
    melt(id.vars = 'Geneid', 
         variable.name = 'stage', 
         value.name = 'tpm') %>% 
  mutate(clade = pir_clade$clade[match(Geneid, 
                                       pir_clade$Geneid)]
         ) %>% 
  group_by(clade) %>% 
  mutate(clade = paste0(clade, ' (n = ', n_distinct(Geneid) ,')')) %>% 
    mutate(stack_group = Geneid) %>% 
    mutate(tpm = ifelse(tpm > 1, tpm, 0)) %>% #Filtering step for low value pir genes.
    group_by(stage, 
             stack_group,
             clade) %>% 
    dplyr::summarise(total_tpm = sum(tpm)) %>%
  arrange(
          desc(clade),
          total_tpm
)

@@

## Figure 2B - sub-family proportions

@@

@@

## Gametocyte marker gene expression - supp figure 2
